---
title: "RNA-Seq in HPC - Package installation"
author: "Prasanth Sivakumar"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Package installation guide for Part 2 - RNA-Seq analysis workflow using High Performance Computing

Please make sure you have R (version 4 or newer) and RStudio installed for this course. We will be running RStudio on your own machine, not Rosalind, so choose the version for whichever operating system you have.

[Install R](https://www.r-project.org) \
[Install RStudio](https://www.rstudio.com)

You will require three libraries for running the analyses in this course (in addition to knitr to generate the course guide in html and BiocManager to install DESeq2 if you do not already have it)

1. [**BiocManager**](https://cran.r-project.org/web/packages/BiocManager/): package manager for Bioconductor packages for high-throughput genomic analysis 
1. [**knitr**](https://cran.r-project.org/web/packages/knitr/index.html): package used to generate reports from Rmarkdown scripts
1. [**tidyverse**](https://cran.r-project.org/web/packages/tidyverse/): set of packages with numerous functions useful for data analysis
1. [**DESeq2**](https://bioconductor.org/packages/release/bioc/html/DESeq2.html): differential gene expression analysis package
1. [**pheatmap**](https://cran.r-project.org/web/packages/pheatmap/): package for plotting heatmaps

To check if you have these packages installed, please run:
```{r check_package_installation, warning=FALSE, message=FALSE}
req.packages <- c('knitr', 'BiocManager', 'tidyverse', 'DESeq2', 'pheatmap')
data.frame(installed=sapply(req.packages, require, character.only = TRUE))
```

In the output dataframe, all rows should return TRUE. If any of the package name rows returns FALSE, you will need to install the package. 

If any of tidyverse, knitr, pheatmap or BiocManager return FALSE, install it with:

install.packages('package name')



If DESeq2 returns FALSE, make sure BiocManager is installed first. Then use:

BiocManager::install('DESeq2')



When installing packages, you may see messages about attaching of additional packages, masking of objects and package versions. These do not typically need to be addressed. 

Once you have installed the required packages, please run the check again to confirm. All outputs should return TRUE

```{r check_package_again}
req.packages <- c('knitr', 'BiocManager', 'tidyverse', 'DESeq2', 'pheatmap')
data.frame(installed=sapply(req.packages, require, character.only = TRUE))
```
